\documentclass[runningheads]{llncs}
\pagestyle{plain} % turn on page numbers
\usepackage[T1]{fontenc}
%\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}

\usepackage{mathpartir}
\usepackage{amsfonts}
\usepackage{amsmath,amssymb}

\usepackage{isabelle,isabellesym}
\usepackage{latexsym}
%because of orcidlink
\usepackage{xcolor}
% this should be the last package used
\usepackage{pdfsetup}
%clashes with pdfsetup
\usepackage{orcidlink}

\urlstyle{rm}

%%% Isabelle specific adjustments

\isabellestyle{literal}

\usepackage{etoolbox}% `toggle'

\newtoggle{mathprime}
\newtoggle{SF}

\renewcommand{\isadigit}[1]{\ensuremath{#1}}
\let\oldisadigit\isadigit
\let\oldisacharprime\isacharprime

\renewcommand{\isaconst}[1]{\isa{\toggletrue{SF}\toggletrue{mathprime}\textsf{#1}}}
\renewcommand{\isafree}[1]{{\toggletrue{mathprime}#1}}
\renewcommand{\isabound}[1]{{\toggletrue{mathprime}#1}}
\renewcommand{\isavar}[1]{{\toggletrue{mathprime}#1}}

\renewcommand{\isacharprime}{\iftoggle{mathprime}{\isamath{'}}{\oldisacharprime}}
\renewcommand{\isadigit}[1]{\iftoggle{SF}{\textsf{#1}}{\oldisadigit{#1}}}

\newcommand{\noquotes}[1]{{\renewcommand{\isachardoublequote}{}\renewcommand{\isachardoublequoteopen}{}\renewcommand{\isachardoublequoteclose}{}#1}}

\newlength{\funheadersep}
\setlength{\funheadersep}{\smallskipamount}

\newcommand{\hide}[1]{}

%%% End of Isabelle specifics

% no right margin in quote:
%\renewenvironment{quote}
%{\list{}{}%
%\item\relax}
%{\endlist}
% no margins in quote:
\renewenvironment{quote}%
{\list{}{\leftmargin=0mm\rightmargin=0mm}\item[]}%
{\endlist}

\newcommand{\concept}[1]{\textbf{#1}}

\newcommand{\eqnum}[1]{{\upshape\refstepcounter{equation}\hfill(\theequation\label{#1})}}

\hyphenation{Isa-belle}

\begin{document}
\title{Earley's Recognizer: Implementation and Complexity Analysis in Isabelle/HOL\thanks{Dedicated to Lawrence C. Paulson on the occasion of his 70th birthday}}
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here

\author{Tobias Nipkow\inst{1}\orcidlink{0000-0003-0730-515X} \and
Alexander Haberl\inst{2}}
%
%\authorrunning{T. Nipkow et al.}

\institute{Technical University of Munich, Germany
  \url{https://www.proof.cit.tum.de/~nipkow} \and
  Technical University of Munich, Germany
  \email{alexander.haberl@tum.de}}

\maketitle

\begin{abstract}
The starting point of this paper is a recent formal verification by Nipkow and Rau
of an abstract inductive but not directly executable (i.e.\ recursive) version of Earley's recognizer.
The aim of that work was a particularly simple, abstract formalization and stepwise refinement
towards executability. We continue this work and refine it to an executable
implementation together with a proof of the running time: cubic if the $n$-th element of
a list can be accessed in constant time, and quartic if access takes linear time.
We present the running time analysis with the help of a simple Big $O$ formalization. 
\end{abstract}

\section{Introduction}

Earley's recognizer \cite{Earley-PhD,Earley-CACM} is a famous algorithm for deciding the word problem
for all context-free grammars. It runs in cubic time in general and in quadratic time for unambiguous grammars.
A parser would return one or more parse trees as well.
In this paper we restrict attention to the recognizer.
Earley's recognizer or parser are frequently covered in the literature (e.g.\ \cite{AhoU72})
and are the basis of many further refinements (e.g.\ \cite{Scott-ENTCS08}).
This paper focuses on Earley's original version.

The first rigourous mathematical proof of Earley's algorithm is due to Jones \cite{Jones,Jones:tr}.
This was the starting point of the work by Nipkow and Rau \cite{NipkowR-CBJ24}, our starting point now.
We continues that work as follows.
First we refine the recognizer from sets to lists, then from inductive to recursive, and finally
we augment the lists wih a more efficient array-like data structure.
Then we present a complexity analysis that proves the well-known cubic bound
if we can access the $n$-th element of a list in constant time.
If that access takes linear time, we obtain a quartic bound.
Like our predecessors, we assume that the grammar is $\epsilon$-free.

Earley's algorithm supports parsing as well, but the extension is non-trivial,
at least for non-ambiguous grammars, and Earley's original algorithm could return spurious
parse trees \cite{Tomita}.
Scott \cite[Section 1]{Scott-ENTCS08} gives a short history of the generalizations
of Earley's recognizer to a parser. Part of the difficulty is how to store multiple
parse trees efficiently.
Rau and Nipkow \cite{RauN24} have verified an Earley parser based on the work of Scott.
Their work is more monolithic and aimed less at simple refinement proofs than the strand followed by us.
Their parser correctness theorems consider only a single parse tree.

This paper is largely based on the forthcoming master's thesis by the second author.

%\begin{table}
%\caption{Table captions should be placed above the tables.}\label{tab1}
%\begin{tabular}{|l|l|l|}
%\end{tabular}
%\end{table}

%\begin{theorem}
%\end{theorem}
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
%\begin{proof}
%\end{proof}

\input{Paper}

\end{document}
